---
title: "EPPS Math and Coding Camp"
subtitle: "Calculus"
author: "Instructor: Haien Peng"
format:
  revealjs:
    theme: [serif, custom.scss]
    slide-number: true
    logo: eppslogo.png
    css: logo.css
    scrollable: true
    transition: fade
html-math-method:
  method: mathjax
  url: mathjax2/MathJax.js?config=TeX-AMS_HTML-full
execute:
  echo: false
  warning: false
  message: false
  cache: false
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
par(family = "serif")
```

## Why calculus is the language of economic change

::: fragment
- Economic models ask how an outcome changes when a choice, price,
  policy, or parameter changes.
:::

::: fragment
- A **derivative** measures local change.
:::

::: fragment
- A **total derivative** tracks several channels of change at once.
:::

::: fragment
- An **optimization condition** turns economic behavior into equations
  we can analyze.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 3-9, 22-38.
:::

------------------------------------------------------------------------

## Our path today

$$\text{Marginal change}
\longrightarrow
\text{Multivariable change}
\longrightarrow
\text{Comparative statics}
\longrightarrow
\text{Optimization}$$

::: fragment
By the end, you should be able to **read, compute, and interpret** the
calculus that appears in first-year economics.
:::

::: notes
\[Sources\] - Simon and Blume (1994), Chapters 2-5 and 13-19.
:::

------------------------------------------------------------------------

## Learning objectives

By the end of this session, you should be able to:

::: fragment
1.  interpret derivatives, elasticities, and Taylor approximations;
:::

::: fragment
2.  compute gradients, total differentials, and Hessians;
:::

::: fragment
3.  use implicit differentiation for comparative statics;
:::

::: fragment
4.  solve basic unconstrained and equality-constrained optimization
    problems;
:::

::: fragment
5.  explain the economic meaning of first-order conditions and
    multipliers.
:::

------------------------------------------------------------------------

## Warm-up: what do you already recognize?

Without calculating everything, identify what each object is asking for.

::: fragment
1.  $\displaystyle \frac{d}{dp}\left(Ap^{-\varepsilon}\right)$
:::

::: fragment
2.  $\displaystyle \frac{\partial F(K,L)}{\partial K}$
:::

::: fragment
3.  $\displaystyle F(p,t)=0 \quad\Rightarrow\quad \frac{dp^*}{dt}=?$
:::

::: fragment
4.  $\displaystyle \max_x f(x) \quad \text{s.t.}\quad g(x)=m$
:::

::: notes
Do not grade this. Ask students to name the operation and its economic
purpose. Return to the four expressions near the end.
:::

# 1. Marginal change

## A derivative is a local comparison

For a function $y=f(x)$,

$$f'(x)=\lim_{h\to 0}\frac{f(x+h)-f(x)}{h}.$$

::: fragment
- Numerator: change in the outcome.
:::

::: fragment
- Denominator: change in the input.
:::

::: fragment
- Limit: make the comparison local.
:::

::: fragment
Economically, $f'(x)$ is a **marginal effect**, holding the model fixed.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 22-29.
:::

------------------------------------------------------------------------

## The derivative rules are a compact toolkit

| Function      | Derivative                     |
|---------------|--------------------------------|
| $x^a$         | $ax^{a-1}$                     |
| $af(x)+bg(x)$ | $af'(x)+bg'(x)$                |
| $f(x)g(x)$    | $f'(x)g(x)+f(x)g'(x)$          |
| $f(x)/g(x)$   | $[f'(x)g(x)-f(x)g'(x)]/g(x)^2$ |
| $f(g(x))$     | $f'(g(x))g'(x)$                |
| $e^x$         | $e^x$                          |
| $\log x$      | $1/x$                          |

::: notes
\[Sources\] - Simon and Blume (1994), pp. 25-28, 70-74, 93-96.
:::

------------------------------------------------------------------------

## The chain rule follows the channel of change

Suppose output depends on capital, and capital depends on the interest
rate:

$$Y=F(K), \qquad K=K(r).$$

::: fragment
Then

$$\frac{dY}{dr}
=
\frac{dF}{dK}\frac{dK}{dr}.$$
:::

::: fragment
Interpretation:

$$\text{effect of }r\text{ on }Y
=
\text{effect of }K\text{ on }Y
\times
\text{effect of }r\text{ on }K.$$
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 70-74.
:::

------------------------------------------------------------------------

## Elasticity puts marginal change in percentage units

For $y=f(x)$, the elasticity of $y$ with respect to $x$ is

$$\varepsilon_{y,x}
=
\frac{dy}{dx}\frac{x}{y}
=
\frac{d\log y}{d\log x}.$$

::: fragment
Elasticity answers:

> By approximately what percent does $y$ change when $x$ rises by one
> percent?
:::

::: fragment
It is unit-free, so it supports comparisons across markets and scales.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 64-69, 100-106.
:::

------------------------------------------------------------------------

## Constant-elasticity demand makes the interpretation transparent

Let

$$q(p)=Ap^{-\varepsilon}, \qquad A>0,\ \varepsilon>0.$$

::: fragment
Then

$$q'(p)=-\varepsilon Ap^{-\varepsilon-1}$$

and

$$\frac{dq}{dp}\frac{p}{q}=-\varepsilon.$$
:::

::: fragment
The exponent is the price elasticity, with the sign reflecting the
downward-sloping demand curve.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 64-69, 100-106.
:::

------------------------------------------------------------------------

## R makes the local nature of a derivative visible

```{r}
#| label: fig-demand-tangent
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "88%"
p <- seq(1, 8, length.out = 300)
A <- 100
eps <- 1.4
q <- A * p^(-eps)
p0 <- 3
q0 <- A * p0^(-eps)
slope <- -eps * A * p0^(-eps - 1)
tangent <- q0 + slope * (p - p0)
plot(p, q, type = "l", lwd = 3, col = "#154360",
     xlab = "Price, p", ylab = "Quantity demanded, q(p)",
     main = "Demand curve and its local linear approximation")
lines(p, tangent, lwd = 2, lty = 2, col = "#C0392B")
points(p0, q0, pch = 19, cex = 1.3, col = "#C0392B")
legend("topright", c("q(p)", "tangent at p = 3"),
       col = c("#154360", "#C0392B"), lwd = c(3, 2),
       lty = c(1, 2), bty = "n")
```

::: notes
\[Sources\] - Simon and Blume (1994), pp. 22-38, 64-69.
:::

------------------------------------------------------------------------

## Quick check: elasticity

Suppose $q(p)=400p^{-2}$.

::: fragment
1.  Is demand upward- or downward-sloping?
:::

::: fragment
2.  What is the price elasticity?
:::

::: fragment
3.  If price increases by 1%, what is the approximate percentage change
    in quantity?
:::

::: fragment
**Answer:** elasticity is $-2$, so quantity falls by approximately 2%.
:::

------------------------------------------------------------------------

## Taylor approximation turns a nonlinear model into a local linear model

Near $x_0$,

$$f(x)
\approx
f(x_0)+f'(x_0)(x-x_0).$$

::: fragment
Including curvature gives

$$f(x)
\approx
f(x_0)+f'(x_0)(x-x_0)
+\frac{1}{2}f''(x_0)(x-x_0)^2.$$
:::

::: fragment
Economists use this idea for approximation, log-linearization,
comparative statics, and asymptotic statistics.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 34-38 and 827-835.
:::

------------------------------------------------------------------------

## First-order approximation is accurate locally

```{r}
#| label: fig-taylor
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "88%"
x <- seq(-1.4, 1.4, length.out = 300)
truth <- exp(x)
first <- 1 + x
second <- 1 + x + x^2 / 2
plot(x, truth, type = "l", lwd = 3, col = "#154360",
     xlab = "x", ylab = "Value",
     main = "Taylor approximations to exp(x) around x = 0")
lines(x, first, lwd = 2, lty = 2, col = "#C0392B")
lines(x, second, lwd = 2, lty = 3, col = "#1E8449")
abline(v = 0, col = "grey70")
legend("topleft", c("exp(x)", "first order", "second order"),
       col = c("#154360", "#C0392B", "#1E8449"),
       lwd = c(3, 2, 2), lty = c(1, 2, 3), bty = "n")
```

::: notes
\[Sources\] - Simon and Blume (1994), pp. 34-38 and 827-835.
:::

# 2. Multivariable change

## Most economic outcomes depend on several inputs

Consider a Cobb-Douglas production function:

$$Y=F(K,L)=AK^\alpha L^\beta.$$

::: fragment
The marginal product of capital is

$$F_K(K,L)=\frac{\partial F}{\partial K}
=\alpha AK^{\alpha-1}L^\beta.$$
:::

::: fragment
The partial derivative changes $K$ while holding $L$ fixed.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 273-299, 300-305.
:::

------------------------------------------------------------------------

## The gradient collects all first-order marginal effects

For $f:\mathbb R^n\rightarrow\mathbb R$,

$$\nabla f(x)
=
\begin{bmatrix}
\partial f/\partial x_1\\
\vdots\\
\partial f/\partial x_n
\end{bmatrix}.$$

::: fragment
For $F(K,L)$,

$$\nabla F(K,L)
=
\begin{bmatrix}
F_K(K,L)\\
F_L(K,L)
\end{bmatrix}.$$
:::

::: fragment
The gradient points in the direction of the steepest local increase.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 319-322.
:::

------------------------------------------------------------------------

## The total differential adds the channels of change

For $z=f(x,y)$,

$$dz=f_x(x,y)\,dx+f_y(x,y)\,dy.$$

::: fragment
In vector notation,

$$df=\nabla f(x)'dx.$$
:::

::: fragment
For production,

$$dY\approx F_K\,dK+F_L\,dL.$$
:::

::: fragment
This decomposes the change in output into a capital channel and a labor
channel.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 307-312, 323-327.
:::

------------------------------------------------------------------------

## Multivariable chain rule: direct and indirect effects

Suppose utility depends on consumption, while consumption depends on
income:

$$U=U(c_1,c_2),
\qquad
c_1=c_1(m),\quad c_2=c_2(m).$$

::: fragment
Then

$$\frac{dU}{dm}
=
U_{c_1}\frac{dc_1}{dm}
+
U_{c_2}\frac{dc_2}{dm}.$$
:::

::: fragment
Every channel contributes a marginal effect multiplied by the response
of that channel.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 313-318, 323-327.
:::

------------------------------------------------------------------------

## The Hessian collects curvature

For $f(x_1,\ldots,x_n)$,

$$H_f(x)=
\begin{bmatrix}
f_{11} & \cdots & f_{1n}\\
\vdots & \ddots & \vdots\\
f_{n1} & \cdots & f_{nn}
\end{bmatrix}.$$

::: fragment
The second-order approximation is

$$f(x+\Delta x)
\approx
f(x)+\nabla f(x)'\Delta x
+\frac{1}{2}\Delta x'H_f(x)\Delta x.$$
:::

::: fragment
The gradient gives slope; the Hessian gives curvature and interactions.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 328-333, 832-840.
:::

------------------------------------------------------------------------

## Concavity connects curvature to global behavior

::: fragment
- If $H_f(x)$ is negative definite, $f$ is locally strictly concave at
  $x$.
:::

::: fragment
- If $f$ is concave on its domain, any point satisfying the appropriate
  first-order condition is a global maximizer.
:::

::: fragment
For a $2\times2$ symmetric Hessian,

$$H=
\begin{bmatrix}
f_{11}&f_{12}\\
f_{12}&f_{22}
\end{bmatrix},$$

negative definiteness requires

$$f_{11}<0,
\qquad
\det(H)>0.$$
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 375-385, 398-404, 505-521.
:::

------------------------------------------------------------------------

## Level curves organize multivariable information

```{r}
#| label: fig-level-curves
#| fig-width: 8.5
#| fig-height: 5.2
#| out-width: "82%"
k <- seq(0.4, 10, length.out = 120)
l <- seq(0.4, 10, length.out = 120)
z <- outer(k, l, function(K, L) K^0.4 * L^0.6)
contour(k, l, z, nlevels = 8, lwd = 2, col = "#154360",
        xlab = "Capital, K", ylab = "Labor, L",
        main = "Isoquants of a Cobb-Douglas production function")
points(4, 6, pch = 19, col = "#C0392B", cex = 1.2)
arrows(4, 6, 5.1, 6.5, length = 0.10, lwd = 2, col = "#C0392B")
text(5.4, 6.7, "gradient", col = "#C0392B", pos = 4)
```

::: notes
\[Sources\] - Simon and Blume (1994), pp. 277-286, 319-322, 342-349.
:::

------------------------------------------------------------------------

## Quick check: total change

Let $Y=K^{1/3}L^{2/3}$.

::: fragment
1.  Find $Y_K$ and $Y_L$.
:::

::: fragment
2.  Write the total differential $dY$.
:::

::: fragment
3.  If both inputs rise by 1%, what is the approximate percentage change
    in output?
:::

::: fragment
**Answer:** approximately 1%, because the output elasticities sum to
one.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 302-312, 483-492.
:::

# 3. Comparative statics

## Equilibrium is often defined by an equation, not a closed form

Suppose an equilibrium $y^*$ satisfies

$$F(y^*,\theta)=0,$$

where $\theta$ is an exogenous parameter.

::: fragment
The comparative-static question is

$$\frac{dy^*}{d\theta}=?$$
:::

::: fragment
We do not always need to solve explicitly for $y^*(\theta)$.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 334-341, 360-363.
:::

------------------------------------------------------------------------

## Implicit differentiation gives the local response

Differentiate the equilibrium condition:

$$F_y\,dy^*+F_\theta\,d\theta=0.$$

::: fragment
If $F_y\neq0$, then

$$\boxed{
\frac{dy^*}{d\theta}
=
-\frac{F_\theta}{F_y}
}$$
:::

::: fragment
The denominator condition is what allows the equilibrium relation to be
locally solved for $y$.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 334-341.
:::

------------------------------------------------------------------------

## Example: a per-unit tax shifts market equilibrium

Let demand and supply be

$$Q_D=a-bp,
\qquad
Q_S=c+d(p-t).$$

Equilibrium satisfies

$$F(p,t)=a-bp-c-d(p-t)=0.$$

::: fragment
Therefore,

$$\frac{dp^*}{dt}
=
-\frac{F_t}{F_p}
=
\frac{d}{b+d}>0.$$
:::

::: fragment
The consumer price rises, but generally by less than the full tax.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 360-363; the linear
supply-demand application is instructor-developed.
:::

------------------------------------------------------------------------

## R shows the equilibrium before and after the tax

```{r}
#| label: fig-tax-equilibrium
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "88%"
q <- seq(0, 90, length.out = 300)
a <- 100; b <- 2; c0 <- 10; d <- 1.5; tax <- 12
pd <- (a - q) / b
ps0 <- (q - c0) / d
ps1 <- tax + ps0
plot(q, pd, type = "l", lwd = 3, col = "#154360",
     xlab = "Quantity", ylab = "Price",
     ylim = range(c(pd, ps0, ps1)),
     main = "A per-unit tax shifts the supply curve upward")
lines(q, ps0, lwd = 2, col = "#1E8449")
lines(q, ps1, lwd = 2, lty = 2, col = "#C0392B")
legend("topright", c("Demand", "Supply", "Supply + tax"),
       col = c("#154360", "#1E8449", "#C0392B"),
       lwd = c(3, 2, 2), lty = c(1, 1, 2), bty = "n")
```

::: notes
\[Sources\] - Simon and Blume (1994), pp. 360-363; figure and
calibration are instructor-developed.
:::

------------------------------------------------------------------------

## Comparative statics is about signs, magnitudes, and conditions

When you derive $dy^*/d\theta$, ask:

::: fragment
1.  **Sign:** does the equilibrium rise or fall?
:::

::: fragment
2.  **Magnitude:** how sensitive is the equilibrium?
:::

::: fragment
3.  **Mechanism:** which derivative drives the result?
:::

::: fragment
4.  **Condition:** what must be nonzero or have a particular sign?
:::

::: fragment
5.  **Scope:** is the result local or global?
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 334-363.
:::

# 4. Optimization

## Optimization turns behavior into equations

An unconstrained choice problem has the form

$$\max_x f(x).$$

::: fragment
At an interior optimum $x^*$, a small feasible change cannot improve the
objective.
:::

::: fragment
This intuition produces the first-order condition

$$f'(x^*)=0.$$
:::

::: fragment
But a zero derivative identifies a candidate, not automatically a
maximum.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 51-57, 396-404.
:::

------------------------------------------------------------------------

## First-order and second-order conditions do different jobs

For a one-dimensional interior optimum:

::: fragment
- **FOC:** $f'(x^*)=0$ identifies stationary points.
:::

::: fragment
- **SOC for a strict local maximum:** $f''(x^*)<0$ verifies downward
  curvature.
:::

::: fragment
- **Global statement:** concavity ensures a stationary point is a global
  maximum.
:::

::: fragment
Boundary points must be checked separately.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 51-57, 396-404, 505-521.
:::

------------------------------------------------------------------------

## Example: a profit-maximizing firm sets marginal revenue equal to marginal cost

Let

$$\pi(q)=R(q)-C(q).$$

::: fragment
The first-order condition is

$$\pi'(q^*)=R'(q^*)-C'(q^*)=0,$$

so

$$MR(q^*)=MC(q^*).$$
:::

::: fragment
The economic rule is the mathematical condition with labels attached.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 58-64, 404-410.
:::

------------------------------------------------------------------------

## A graph separates the candidate from the curvature condition

```{r}
#| label: fig-profit-max
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "88%"
q <- seq(0, 24, length.out = 300)
price <- 30
cost <- 40 + 6 * q + 0.6 * q^2
profit <- price * q - cost
qstar <- (price - 6) / 1.2
pistar <- price * qstar - (40 + 6 * qstar + 0.6 * qstar^2)
plot(q, profit, type = "l", lwd = 3, col = "#154360",
     xlab = "Output, q", ylab = "Profit, pi(q)",
     main = "The FOC locates the peak; curvature verifies it")
abline(h = 0, col = "grey75")
points(qstar, pistar, pch = 19, cex = 1.3, col = "#C0392B")
abline(v = qstar, lty = 2, col = "#C0392B")
```

::: notes
\[Sources\] - Simon and Blume (1994), pp. 58-64, 396-410; calibration is
instructor-developed.
:::

------------------------------------------------------------------------

## In several dimensions, the gradient must vanish

For

$$\max_{x\in\mathbb R^n} f(x),$$

an interior optimum satisfies

$$\nabla f(x^*)=0.$$

::: fragment
For two variables, this is a system:

$$f_{x_1}(x_1^*,x_2^*)=0,
\qquad
f_{x_2}(x_1^*,x_2^*)=0.$$
:::

::: fragment
The Hessian then provides the local curvature test.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 396-404.
:::

------------------------------------------------------------------------

## A constraint changes which directions are feasible

Consumer choice has the form

$$\max_{x_1,x_2} u(x_1,x_2)
\quad\text{s.t.}\quad
p_1x_1+p_2x_2=m.$$

::: fragment
At an interior optimum, the highest attainable indifference curve is
tangent to the budget line.
:::

::: fragment
The constraint means we cannot perturb $x_1$ and $x_2$ independently.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 5-9, 232-233, 411-423,
544-556.
:::

------------------------------------------------------------------------

## The Lagrangian incorporates the constraint

Write

$$\mathcal L(x_1,x_2,\lambda)
=
u(x_1,x_2)
+\lambda(m-p_1x_1-p_2x_2).$$

::: fragment
The first-order conditions are

$$u_{x_1}-\lambda p_1=0,
\qquad
u_{x_2}-\lambda p_2=0,
\qquad
m-p_1x_1-p_2x_2=0.$$
:::

::: fragment
These equations jointly determine $x_1^*$, $x_2^*$, and $\lambda^*$.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 411-423.
:::

------------------------------------------------------------------------

## Tangency equates marginal benefit per dollar

Dividing the first two conditions gives

$$\frac{u_{x_1}}{u_{x_2}}
=
\frac{p_1}{p_2}.$$

Equivalently,

$$\frac{u_{x_1}}{p_1}
=
\frac{u_{x_2}}{p_2}
=\lambda.$$

::: fragment
At an interior optimum, the last dollar spent on either good generates
the same marginal utility.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 411-423, 544-556.
:::

------------------------------------------------------------------------

## Worked example: Cobb-Douglas demand

Let

$$u(x_1,x_2)=x_1^\alpha x_2^{1-\alpha},
\qquad 0<\alpha<1.$$

::: fragment
Solving the FOCs and the budget constraint gives

$$x_1^*=\frac{\alpha m}{p_1},
\qquad
x_2^*=\frac{(1-\alpha)m}{p_2}.$$
:::

::: fragment
The consumer spends a constant share $\alpha$ on good 1 and $1-\alpha$
on good 2.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 411-423, 544-556.
:::

------------------------------------------------------------------------

## The multiplier is a shadow value

Let the value function be

$$V(m)=\max_x f(x)
\quad\text{s.t.}\quad g(x)=m.$$

::: fragment
Under regularity conditions,

$$\frac{dV(m)}{dm}=\lambda^*.$$
:::

::: fragment
$\lambda^*$ measures how much the optimized objective changes when the
constraint is relaxed by one unit.
:::

::: fragment
Examples: marginal utility of income, value of capacity, or value of a
policy resource.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 448-456.
:::

------------------------------------------------------------------------

## The envelope theorem avoids differentiating the optimizer

Suppose

$$V(\theta)=\max_x f(x,\theta).$$

At an interior optimum,

$$\frac{dV}{d\theta}
=
f_x(x^*,\theta)\frac{dx^*}{d\theta}
+f_\theta(x^*,\theta).$$

::: fragment
Because $f_x(x^*,\theta)=0$,

$$\boxed{\frac{dV}{d\theta}=f_\theta(x^*,\theta)}.$$
:::

::: fragment
At the optimum, the indirect effect through the optimizing choice
vanishes to first order.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 453-456.
:::

------------------------------------------------------------------------

## One calculus toolkit appears across the PhD core

| Tool | Microeconomics | Macroeconomics | Econometrics |
|------------------|------------------|------------------|------------------|
| Derivative | marginal utility | growth rates | score function |
| Gradient | FOCs | equilibrium systems | normal equations |
| Hessian | concavity | local dynamics | information matrix |
| IFT | demand responses | policy comparative statics | estimator sensitivity |
| Taylor expansion | welfare approximation | log-linearization | delta method |
| Optimization | consumer/firm choice | planner problem | MLE/GMM |

::: notes
\[Sources\] - Simon and Blume (1994), Chapters 14-19. - Hansen (2021),
pp. 170-173, 192-217.
:::

------------------------------------------------------------------------

## Common mistakes to catch early

::: fragment
- Treating a partial derivative as if every other variable also changes.
:::

::: fragment
- Forgetting the inner derivative in the chain rule.
:::

::: fragment
- Treating a first-order condition as proof of a maximum.
:::

::: fragment
- Ignoring boundary or corner solutions.
:::

::: fragment
- Interpreting a local comparative-static result as a global theorem.
:::

::: fragment
- Reporting a derivative without units or economic interpretation.
:::

------------------------------------------------------------------------

## Practice: connect the tools

Consider a competitive firm with production $q=\sqrt{L}$, output price
$p$, and wage $w$.

$$\pi(L)=p\sqrt L-wL.$$

::: fragment
1.  Derive the firm's labor demand $L^*(p,w)$.
:::

::: fragment
2.  Determine the sign of $\partial L^*/\partial w$.
:::

::: fragment
3.  Find the optimized profit $\pi^*(p,w)$.
:::

::: fragment
4.  Use the envelope theorem to find $\partial\pi^*/\partial w$.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 396-410, 453-456, 557-564.
:::

------------------------------------------------------------------------

## Practice solution

The FOC is

$$\frac{p}{2\sqrt L}-w=0,$$

so

$$L^*(p,w)=\frac{p^2}{4w^2}.$$

::: fragment
Therefore,

$$\frac{\partial L^*}{\partial w}
=-\frac{p^2}{2w^3}<0.$$
:::

::: fragment
The optimized profit is

$$\pi^*(p,w)=\frac{p^2}{4w}.$$
:::

------------------------------------------------------------------------

## The envelope check

Direct differentiation gives

$$\frac{\partial\pi^*}{\partial w}
=-\frac{p^2}{4w^2}.$$

::: fragment
The envelope theorem gives the same answer immediately:

$$\frac{\partial\pi^*}{\partial w}
=
\left.
\frac{\partial\pi(L;p,w)}{\partial w}
\right|_{L=L^*}
=-L^*
=-\frac{p^2}{4w^2}.$$
:::

::: fragment
Economic interpretation: a marginal wage increase reduces optimized
profit by the firm's optimal labor demand.
:::

::: notes
\[Sources\] - Simon and Blume (1994), pp. 453-456, 557-564.
:::

------------------------------------------------------------------------

## Exit ticket

In one sentence or one equation, answer each question.

::: fragment
1.  What is the difference between a partial derivative and a total
    derivative?
:::

::: fragment
2.  What additional condition turns an FOC candidate into a maximum?
:::

::: fragment
3.  What does $-F_\theta/F_y$ measure?
:::

::: fragment
4.  What is the economic meaning of a Lagrange multiplier?
:::

------------------------------------------------------------------------

## Keep this mental map

$$\boxed{
\begin{array}{c}
\text{Derivative}=\text{marginal change}\\[3pt]
\text{Gradient}=\text{all marginal changes}\\[3pt]
\text{Hessian}=\text{curvature}\\[3pt]
\text{IFT}=\text{equilibrium response}\\[3pt]
\text{FOC + curvature}=\text{optimal choice}
\end{array}}$$

::: fragment
The goal is not to memorize isolated formulas. It is to recognize which
economic question each object answers.
:::

------------------------------------------------------------------------

**Questions?**

Primary reference: Simon and Blume, *Mathematics for Economists*,
Chapters 2-5 and 13-19.
