---
title: "EPPS Math and Coding Camp"
subtitle: "Probability and Statistics"
author: "Instructor: Haien Peng"
format:
  revealjs:
    theme: [serif, custom.scss]
    slide-number: true
    logo: eppslogo.png
    css: logo.css
    scrollable: true
    transition: fade
html-math-method:
  method: mathjax
  url: mathjax2/MathJax.js?config=TeX-AMS_HTML-full
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

```{r}
#| label: setup
#| include: false
set.seed(2026)
par(family = "serif")
```

## Statistics connects an unseen population to observed data

$$
\text{Population}
\longrightarrow
\text{Random sample}
\longrightarrow
\text{Estimator}
\longrightarrow
\text{Uncertainty}
\longrightarrow
\text{Inference}
$$

::: fragment
Probability describes what data could be generated.
:::

::: fragment
Statistics uses generated data to learn about the process.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 1-17, 128-140.
:::

------------------------------------------------------------------------

## Our path today

::: fragment
1. Probability: how uncertainty is represented.
:::

::: fragment
2. Random variables: how outcomes become numerical objects.
:::

::: fragment
3. Sampling: why estimates vary across samples.
:::

::: fragment
4. LLN and CLT: why large samples help.
:::

::: fragment
5. Estimation and inference: what we can conclude from data.
:::

------------------------------------------------------------------------

## Learning objectives

By the end of this session, you should be able to:

::: fragment
- distinguish a population parameter from a sample statistic;
:::

::: fragment
- compute and interpret probability, expectation, variance, and covariance;
:::

::: fragment
- explain conditional expectation, the LLN, and the CLT;
:::

::: fragment
- distinguish standard deviation from standard error;
:::

::: fragment
- interpret a confidence interval, test statistic, and p-value correctly.
:::

------------------------------------------------------------------------

## Warm-up: which statements are correct?

::: fragment
1. A parameter changes every time we draw a new sample.
:::

::: fragment
2. A 5% p-value means the null hypothesis has a 5% probability of being true.
:::

::: fragment
3. A larger sample always eliminates bias.
:::

::: fragment
4. If $X$ and $Y$ are uncorrelated, they must be independent.
:::

::: fragment
**Preview:** all four statements are generally false.
:::

# 1. Probability

## Outcomes, events, and sample spaces

::: fragment
- An **outcome** is one possible realization.
:::

::: fragment
- The **sample space** $\Omega$ contains all possible outcomes.
:::

::: fragment
- An **event** $A$ is a subset of $\Omega$.
:::

::: fragment
For a binary employment outcome,

$$
\Omega=\{0,1\},
\qquad
A=\{\text{employed}\}=\{1\}.
$$
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 1-4.
:::

------------------------------------------------------------------------

## Probability obeys three basic rules

For events in $\Omega$:

::: fragment
1. $P(A)\geq0$.
:::

::: fragment
2. $P(\Omega)=1$.
:::

::: fragment
3. If $A_1,A_2,\ldots$ are mutually exclusive,

$$
P\left(\bigcup_j A_j\right)=\sum_jP(A_j).
$$
:::

::: fragment
These axioms generate the other probability rules we use.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 3-5.
:::

------------------------------------------------------------------------

## “Or” requires subtracting the overlap

For any two events,

$$
P(A\cup B)
=P(A)+P(B)-P(A\cap B).
$$

::: fragment
Why subtract?
:::

::: fragment
The intersection $A\cap B$ was counted once inside $P(A)$ and once inside $P(B)$.
:::

::: fragment
If $A$ and $B$ are mutually exclusive, then $P(A\cap B)=0$.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 4-6.
:::

------------------------------------------------------------------------

## Conditional probability changes the reference population

For $P(B)>0$,

$$
P(A\mid B)=\frac{P(A\cap B)}{P(B)}.
$$

::: fragment
The denominator says: restrict attention to cases where $B$ occurred.
:::

::: fragment
The numerator says: among those cases, count where $A$ also occurred.
:::

::: fragment
In general,

$$
P(A\mid B)\neq P(B\mid A).
$$
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 6-7.
:::

------------------------------------------------------------------------

## Independence and mutual exclusivity are different ideas

Events $A$ and $B$ are independent if

$$
P(A\cap B)=P(A)P(B).
$$

Equivalently, when probabilities are positive,

$$
P(A\mid B)=P(A).
$$

::: fragment
- **Independent:** learning $B$ does not change the probability of $A$.
:::

::: fragment
- **Mutually exclusive:** $A$ and $B$ cannot occur together.
:::

::: fragment
Nontrivial mutually exclusive events are therefore not independent.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 5-9.
:::

------------------------------------------------------------------------

## The law of total probability decomposes an event

If $B_1,\ldots,B_K$ form a partition of $\Omega$, then

$$
P(A)=\sum_{k=1}^K P(A\mid B_k)P(B_k).
$$

::: fragment
For a treatment indicator $D\in\{0,1\}$,

$$
P(Y=1)
=P(Y=1\mid D=1)P(D=1)
+P(Y=1\mid D=0)P(D=0).
$$
:::

::: fragment
Overall probability is a weighted average of group-specific probabilities.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 9-10.
:::

------------------------------------------------------------------------

## Bayes' rule reverses the conditioning direction

$$
P(B\mid A)
=
\frac{P(A\mid B)P(B)}{P(A)}.
$$

Using the total probability rule,

$$
P(B\mid A)
=
\frac{P(A\mid B)P(B)}
{P(A\mid B)P(B)+P(A\mid B^c)P(B^c)}.
$$

::: fragment
Bayes' rule updates a prior probability after observing evidence.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 9-11.
:::

------------------------------------------------------------------------

## Example: selection into a training program

Suppose:

$$
P(D=1)=0.20,
$$

$$
P(\text{high motivation}\mid D=1)=0.70,
$$

$$
P(\text{high motivation}\mid D=0)=0.25.
$$

::: fragment
What is

$$
P(D=1\mid \text{high motivation})?
$$
:::

::: fragment
First compute the probability of high motivation using the total probability rule.
:::

------------------------------------------------------------------------

## Bayes' rule gives the updated probability

$$
P(H)=0.70(0.20)+0.25(0.80)=0.34.
$$

Therefore,

$$
P(D=1\mid H)
=
\frac{0.70(0.20)}{0.34}
\approx0.412.
$$

::: fragment
Observing high motivation raises the probability of training participation from 20% to about 41%.
:::

::: fragment
This is also a warning: program participation may be statistically related to unobserved motivation.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 9-11; numerical example is instructor-developed.
:::

------------------------------------------------------------------------

## Quick check: conditional probability

In a population, 10% receive a policy treatment. Among treated units, 60% have outcome $Y=1$; among untreated units, 20% have $Y=1$.

::: fragment
1. Find $P(Y=1)$.
:::

::: fragment
2. Find $P(D=1\mid Y=1)$.
:::

::: fragment
**Answer:**

$$
P(Y=1)=0.6(0.1)+0.2(0.9)=0.24,
$$

$$
P(D=1\mid Y=1)=0.06/0.24=0.25.
$$
:::

# 2. Random variables and distributions

## A random variable maps outcomes into numbers

A random variable is a function

$$
X:\Omega\rightarrow\mathbb R.
$$

::: fragment
- Discrete example: number of job offers.
:::

::: fragment
- Continuous example: wage or income.
:::

::: fragment
- Binary example: employment status.
:::

::: fragment
The randomness comes from the outcome; the mapping itself is fixed.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 22-24, 28-34.
:::

------------------------------------------------------------------------

## PMF, PDF, and CDF answer different questions

::: fragment
- Discrete probability mass function:

$$p(x)=P(X=x).$$
:::

::: fragment
- Continuous probability density function:

$$P(a<X\le b)=\int_a^b f(x)\,dx.$$
:::

::: fragment
- Cumulative distribution function:

$$F(x)=P(X\le x).$$
:::

::: fragment
A density value is not itself a probability; probability is area under the density.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 22-39.
:::

------------------------------------------------------------------------

## Expectation is a probability-weighted average

For a discrete random variable,

$$
E[X]=\sum_x xP(X=x).
$$

For a continuous random variable,

$$
E[X]=\int x f(x)\,dx.
$$

::: fragment
Expectation is a feature of the distribution, not necessarily a value that will ever be observed.
:::

::: fragment
Linearity:

$$
E[a+bX+cY]=a+bE[X]+cE[Y].
$$
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 25-27, 37-42.
:::

------------------------------------------------------------------------

## Variance measures dispersion around the mean

Let $\mu=E[X]$. Then

$$
\operatorname{Var}(X)
=E[(X-\mu)^2]
=E[X^2]-\mu^2.
$$

::: fragment
The standard deviation is

$$
\operatorname{SD}(X)=\sqrt{\operatorname{Var}(X)}.
$$
:::

::: fragment
Standard deviation uses the same units as $X$; variance uses squared units.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 40-42.
:::

------------------------------------------------------------------------

## Three distributions appear repeatedly

::: fragment
**Bernoulli:** $X\in\{0,1\}$ and $P(X=1)=p$.

$$E[X]=p,\qquad \operatorname{Var}(X)=p(1-p).$$
:::

::: fragment
**Binomial:** number of successes in $n$ independent Bernoulli trials.
:::

::: fragment
**Normal:** continuous, symmetric, and described by mean $\mu$ and variance $\sigma^2$.
:::

::: fragment
Student $t$, $\chi^2$, and $F$ distributions enter classical inference.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 56-64, 113-126.
:::

------------------------------------------------------------------------

## R connects the formulas to distributional shapes

```{r}
#| label: fig-distributions
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "90%"
old <- par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
k <- 0:12
plot(k, dbinom(k, size = 12, prob = 0.35), type = "h", lwd = 5,
     col = "#154360", xlab = "Number of successes",
     ylab = "Probability", main = "Binomial(12, 0.35)")
points(k, dbinom(k, 12, 0.35), pch = 19, col = "#154360")
x <- seq(-4, 4, length.out = 400)
plot(x, dnorm(x), type = "l", lwd = 3, col = "#C0392B",
     xlab = "x", ylab = "Density", main = "Standard normal")
polygon(c(x, rev(x)), c(dnorm(x), rep(0, length(x))),
        col = adjustcolor("#C0392B", 0.18), border = NA)
lines(x, dnorm(x), lwd = 3, col = "#C0392B")
par(old)
```

::: {.notes}
[Sources]
- Hansen (2021), pp. 56-64, 113-126.
:::

------------------------------------------------------------------------

## Joint distributions describe variables together

For two random variables $X$ and $Y$, the joint distribution describes

$$
P(X\in A,\ Y\in B).
$$

::: fragment
- A **marginal distribution** focuses on one variable.
:::

::: fragment
- A **conditional distribution** focuses on one variable given the other.
:::

::: fragment
- Independence means the joint distribution factors into the product of marginals.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 74-90.
:::

------------------------------------------------------------------------

## Covariance and correlation summarize co-movement

$$
\operatorname{Cov}(X,Y)
=E[(X-E[X])(Y-E[Y])].
$$

$$
\operatorname{Corr}(X,Y)
=
\frac{\operatorname{Cov}(X,Y)}
{\operatorname{SD}(X)\operatorname{SD}(Y)}.
$$

::: fragment
- Correlation lies between $-1$ and $1$.
:::

::: fragment
- Independence implies zero covariance when moments exist.
:::

::: fragment
- Zero covariance does not generally imply independence.
:::

::: fragment
- Neither covariance nor correlation establishes causality.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 91-94.
:::

------------------------------------------------------------------------

## Conditional expectation is the language of econometrics

$$
m(x)=E[Y\mid X=x].
$$

::: fragment
$m(x)$ is the mean of the conditional distribution of $Y$ given $X=x$.
:::

::: fragment
The decomposition

$$
Y=E[Y\mid X]+\varepsilon
$$

implies

$$
E[\varepsilon\mid X]=0.
$$
:::

::: fragment
Regression methods approximate or estimate conditional expectations.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 94-99.
:::

------------------------------------------------------------------------

## The law of iterated expectations aggregates conditional means

$$
E[Y]=E\left[E[Y\mid X]\right].
$$

::: fragment
Interpretation:

1. compute the mean of $Y$ within each value or group of $X$;
2. average those conditional means over the distribution of $X$.
:::

::: fragment
This is the expectation analogue of the law of total probability.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 94-97.
:::

------------------------------------------------------------------------

## Jensen's inequality links curvature and risk

If $u$ is concave, then

$$
E[u(X)]\le u(E[X]).
$$

::: fragment
A risk-averse decision maker prefers the utility of a certain mean payoff to the expected utility of a risky payoff with that mean.
:::

::: fragment
Curvature, not merely variance, determines how uncertainty affects welfare.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 42-45.
:::

# 3. Sampling and estimators

## A population parameter is fixed; a statistic is random

Suppose $X_1,\ldots,X_n$ are a random sample from a population.

::: fragment
- **Parameter:** an unknown feature such as $\mu=E[X]$.
:::

::: fragment
- **Statistic:** a function of the sample, such as

$$
\bar X=\frac{1}{n}\sum_{i=1}^nX_i.
$$
:::

::: fragment
- **Estimator:** the rule $\bar X$ used to estimate $\mu$.
:::

::: fragment
- **Estimate:** the numerical value obtained in one sample.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 128-135.
:::

------------------------------------------------------------------------

## The sampling distribution describes repeated-sample variation

Imagine repeatedly drawing samples of size $n$ and computing $\bar X$.

::: fragment
The distribution of those values is the **sampling distribution** of $\bar X$.
:::

::: fragment
Under IID sampling with finite variance,

$$
E[\bar X]=\mu,
\qquad
\operatorname{Var}(\bar X)=\frac{\sigma^2}{n}.
$$
:::

::: fragment
More data reduces variance, but the estimate remains random.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 128-140.
:::

------------------------------------------------------------------------

## Standard deviation and standard error answer different questions

::: fragment
**Standard deviation:** how dispersed are individual observations?

$$
\operatorname{SD}(X)=\sigma.
$$
:::

::: fragment
**Standard error:** how dispersed is an estimator across repeated samples?

$$
\operatorname{SE}(\bar X)=\frac{\sigma}{\sqrt n}.
$$
:::

::: fragment
If sample size quadruples, the standard error is approximately halved.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 135-141.
:::

------------------------------------------------------------------------

## R simulates a sampling distribution

```{r}
#| label: fig-sampling-distribution
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "88%"
B <- 5000
n <- 40
means <- replicate(B, mean(rexp(n, rate = 1)))
hist(means, breaks = 35, probability = TRUE,
     col = adjustcolor("#154360", 0.35), border = "white",
     xlab = "Sample mean", main = "Sampling distribution of the mean")
curve(dnorm(x, mean = 1, sd = 1 / sqrt(n)), add = TRUE,
      lwd = 3, col = "#C0392B")
abline(v = 1, lty = 2, lwd = 2, col = "#1E8449")
legend("topright", c("simulated means", "normal approximation", "population mean"),
       col = c("#154360", "#C0392B", "#1E8449"),
       lwd = c(6, 3, 2), lty = c(1, 1, 2), bty = "n")
```

::: {.notes}
[Sources]
- Hansen (2021), pp. 128-140, 165-173.
:::

------------------------------------------------------------------------

## Bias, variance, and MSE evaluate an estimator

Bias:

$$
\operatorname{Bias}(\hat\theta)
=E[\hat\theta]-\theta.
$$

Mean squared error:

$$
\operatorname{MSE}(\hat\theta)
=E[(\hat\theta-\theta)^2]
=\operatorname{Var}(\hat\theta)
+\operatorname{Bias}(\hat\theta)^2.
$$

::: fragment
An estimator can trade a small amount of bias for a large reduction in variance.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 135-140.
:::

# 4. Why large samples help

## The law of large numbers gives consistency

Under suitable conditions,

$$
\bar X_n\xrightarrow{p}\mu.
$$

::: fragment
For every $\varepsilon>0$,

$$
P(|\bar X_n-\mu|>\varepsilon)\rightarrow0.
$$
:::

::: fragment
The LLN says the sample mean becomes close to the population mean in probability.
:::

::: fragment
It does not describe the shape of the remaining error.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 149-160.
:::

------------------------------------------------------------------------

## R shows convergence as sample size grows

```{r}
#| label: fig-lln
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "88%"
nmax <- 1200
x <- rexp(nmax, rate = 1)
running_mean <- cumsum(x) / seq_along(x)
plot(seq_along(x), running_mean, type = "l", lwd = 2,
     col = "#154360", xlab = "Sample size, n",
     ylab = "Running sample mean",
     main = "The running mean settles near the population mean")
abline(h = 1, lwd = 2, lty = 2, col = "#C0392B")
legend("topright", c("running mean", "E[X] = 1"),
       col = c("#154360", "#C0392B"), lwd = 2,
       lty = c(1, 2), bty = "n")
```

::: {.notes}
[Sources]
- Hansen (2021), pp. 149-160.
:::

------------------------------------------------------------------------

## The central limit theorem gives the shape of estimation error

Under suitable conditions,

$$
\sqrt n(\bar X_n-\mu)
\xrightarrow{d}
N(0,\sigma^2).
$$

Equivalently, for large $n$,

$$
\bar X_n
\approx
N\left(\mu,\frac{\sigma^2}{n}\right).
$$

::: fragment
This approximation is the bridge from estimation to standard errors, confidence intervals, and tests.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 165-173.
:::

------------------------------------------------------------------------

## The original population need not be normal

```{r}
#| label: fig-clt
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "90%"
old <- par(mfrow = c(1, 3), mar = c(4, 3.5, 3, 1))
for (n in c(1, 5, 40)) {
  vals <- replicate(4000, mean(rexp(n, rate = 1)))
  hist(vals, breaks = 30, probability = TRUE,
       col = adjustcolor("#154360", 0.35), border = "white",
       xlab = "Sample mean", ylab = "Density",
       main = paste0("n = ", n), xlim = c(0, 3))
  curve(dnorm(x, mean = 1, sd = 1 / sqrt(n)), add = TRUE,
        lwd = 2, col = "#C0392B")
}
par(old)
```

::: {.notes}
[Sources]
- Hansen (2021), pp. 165-173.
:::

------------------------------------------------------------------------

## LLN and CLT answer different questions

| Result | Main question | Main conclusion |
|---|---|---|
| LLN | Does the estimator approach the target? | $\bar X_n\to_p\mu$ |
| CLT | What is the approximate distribution of its error? | $\sqrt n(\bar X_n-\mu)\to_d N(0,\sigma^2)$ |

::: fragment
- LLN supports consistency.
:::

::: fragment
- CLT supports approximate inference.
:::

::: fragment
Both rely on assumptions; neither is a license to ignore dependence, heavy tails, or poor sampling design.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 149-176.
:::

# 5. Estimation and inference

## Method of moments matches sample moments to population moments

Suppose a model implies

$$
E[g(X,\theta_0)]=0.
$$

The method of moments chooses $\hat\theta$ so that

$$
\frac{1}{n}\sum_{i=1}^n g(X_i,\hat\theta)=0.
$$

::: fragment
For $E[X]=\mu$, the sample analogue gives

$$
\hat\mu=\bar X.
$$
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 225-241.
:::

------------------------------------------------------------------------

## Maximum likelihood asks which parameter best explains the sample

For IID observations with density or mass function $f(x;\theta)$,

$$
L(\theta)=\prod_{i=1}^n f(X_i;\theta).
$$

The maximum likelihood estimator is

$$
\hat\theta_{MLE}=\arg\max_\theta L(\theta).
$$

::: fragment
We usually maximize the log-likelihood:

$$
\ell(\theta)=\sum_{i=1}^n\log f(X_i;\theta).
$$
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 192-206.
:::

------------------------------------------------------------------------

## Bernoulli MLE: the sample mean estimates a probability

If $X_i\sim\text{Bernoulli}(p)$, then

$$
\ell(p)
=
\sum_{i=1}^n
\left[X_i\log p+(1-X_i)\log(1-p)\right].
$$

::: fragment
The first-order condition yields

$$
\hat p=\frac{1}{n}\sum_{i=1}^nX_i=\bar X.
$$
:::

::: fragment
The MLE is simply the observed fraction of successes.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 192-206.
:::

------------------------------------------------------------------------

## A confidence interval combines estimate and uncertainty

If

$$
\frac{\hat\theta-\theta}{\operatorname{SE}(\hat\theta)}
\approx N(0,1),
$$

an approximate 95% confidence interval is

$$
\boxed{
\hat\theta
\pm
1.96\operatorname{SE}(\hat\theta)
}.
$$

::: fragment
Across repeated samples, approximately 95% of intervals constructed by this procedure contain the true parameter.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 296-301.
:::

------------------------------------------------------------------------

## Confidence intervals vary from sample to sample

```{r}
#| label: fig-confidence-intervals
#| fig-width: 9
#| fig-height: 5.2
#| out-width: "88%"
B <- 30
n <- 50
mu <- 1
est <- se <- numeric(B)
for (b in seq_len(B)) {
  xb <- rexp(n, rate = 1)
  est[b] <- mean(xb)
  se[b] <- sd(xb) / sqrt(n)
}
lo <- est - 1.96 * se
hi <- est + 1.96 * se
cover <- lo <= mu & hi >= mu
plot(est, seq_len(B), pch = 19,
     col = ifelse(cover, "#154360", "#C0392B"),
     xlim = range(c(lo, hi, mu)), xlab = "Confidence interval",
     ylab = "Repeated sample", main = "Most, but not all, 95% intervals cover the truth")
segments(lo, seq_len(B), hi, seq_len(B),
         col = ifelse(cover, "#154360", "#C0392B"), lwd = 2)
abline(v = mu, lty = 2, lwd = 2, col = "#1E8449")
```

::: {.notes}
[Sources]
- Hansen (2021), pp. 296-301.
:::

------------------------------------------------------------------------

## Hypothesis testing begins with a null claim

Example:

$$
H_0:\theta=\theta_0
\qquad\text{versus}\qquad
H_1:\theta\neq\theta_0.
$$

The standardized test statistic is

$$
t=\frac{\hat\theta-\theta_0}{\operatorname{SE}(\hat\theta)}.
$$

::: fragment
Under $H_0$, a large absolute value of $t$ is evidence against the null.
:::

::: fragment
At the 5% level in a two-sided large-sample test, reject when $|t|>1.96$.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 271-290.
:::

------------------------------------------------------------------------

## Type I and Type II errors are asymmetric

| Decision | $H_0$ true | $H_0$ false |
|---|---|---|
| Reject $H_0$ | Type I error | Correct rejection |
| Do not reject $H_0$ | Correct non-rejection | Type II error |

::: fragment
- Significance level $\alpha$: probability of a Type I error under the null.
:::

::: fragment
- Power: probability of rejecting a false null.
:::

::: fragment
Reducing one error probability can increase the other unless information increases.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 271-281.
:::

------------------------------------------------------------------------

## A p-value measures incompatibility with the null

The p-value is the probability, **assuming the null is true**, of observing a test statistic at least as extreme as the one observed.

::: fragment
For a two-sided asymptotic normal test,

$$
p\text{-value}
=2\left[1-\Phi(|t_{obs}|)\right].
$$
:::

::: fragment
It is not $P(H_0\text{ is true}\mid\text{data})$.
:::

::: fragment
A small p-value is evidence against $H_0$ relative to the assumed model.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 282-290.
:::

------------------------------------------------------------------------

## R displays the two-sided p-value

```{r}
#| label: fig-pvalue
#| fig-width: 9
#| fig-height: 4.8
#| out-width: "88%"
x <- seq(-4, 4, length.out = 800)
y <- dnorm(x)
tobs <- 2.1
plot(x, y, type = "l", lwd = 3, col = "#154360",
     xlab = "Test statistic under H0", ylab = "Density",
     main = "Two-sided p-value: probability beyond +/- |t observed|")
xl <- x[x <= -tobs]
xr <- x[x >= tobs]
polygon(c(xl, rev(xl)), c(dnorm(xl), rep(0, length(xl))),
        col = adjustcolor("#C0392B", 0.45), border = NA)
polygon(c(xr, rev(xr)), c(dnorm(xr), rep(0, length(xr))),
        col = adjustcolor("#C0392B", 0.45), border = NA)
lines(x, y, lwd = 3, col = "#154360")
abline(v = c(-tobs, tobs), lty = 2, col = "#C0392B")
```

::: {.notes}
[Sources]
- Hansen (2021), pp. 280-290.
:::

------------------------------------------------------------------------

## Statistical significance is not economic significance

Suppose an estimated policy effect is

$$
\hat\theta=0.002,
\qquad
\operatorname{SE}(\hat\theta)=0.0005.
$$

::: fragment
Then $t=4$, so the estimate is statistically distinguishable from zero.
:::

::: fragment
But whether an effect of $0.002$ matters depends on units, costs, baseline risk, and the decision context.
:::

::: fragment
Always report and interpret the estimate, its uncertainty, and its substantive magnitude.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 289-301; numerical example is instructor-developed.
:::

------------------------------------------------------------------------

## A larger sample does not solve every problem

Larger $n$ often reduces sampling variance, but it does not automatically fix:

::: fragment
- selection bias;
:::

::: fragment
- omitted variables or endogeneity;
:::

::: fragment
- measurement error;
:::

::: fragment
- dependence ignored by the standard error;
:::

::: fragment
- model misspecification;
:::

::: fragment
- a sample that does not represent the target population.
:::

::: {.notes}
[Sources]
- Hansen (2021), pp. 128-176, 213-217.
:::

------------------------------------------------------------------------

## These ideas become the first regression toolkit

In a regression model,

$$
Y_i=X_i'\beta+u_i,
$$

we use the same sequence:

::: fragment
1. define a population target $\beta$;
:::

::: fragment
2. construct an estimator $\hat\beta$;
:::

::: fragment
3. study bias, variance, and consistency;
:::

::: fragment
4. derive an approximate sampling distribution;
:::

::: fragment
5. compute standard errors, intervals, and tests.
:::

::: fragment
Probability is not separate from econometrics; it is its foundation.
:::

------------------------------------------------------------------------

## Practice: one empirical estimate

A sample of $n=400$ workers gives

$$
\bar X=52.0,
\qquad
s=12.0,
$$

where $X$ is weekly hours worked.

::: fragment
1. Estimate the population mean.
:::

::: fragment
2. Compute the standard error.
:::

::: fragment
3. Construct an approximate 95% confidence interval.
:::

::: fragment
4. Test $H_0:\mu=50$ using a two-sided large-sample test.
:::

------------------------------------------------------------------------

## Practice solution

Point estimate:

$$
\hat\mu=52.0.
$$

Standard error:

$$
SE(\bar X)=\frac{12}{\sqrt{400}}=0.6.
$$

::: fragment
Approximate 95% confidence interval:

$$
52\pm1.96(0.6)
=52\pm1.176
=[50.824,53.176].
$$
:::

------------------------------------------------------------------------

## Test and interpretation

$$
t=\frac{52-50}{0.6}=3.33.
$$

::: fragment
The two-sided p-value is approximately

$$
2[1-\Phi(3.33)]\approx0.0009.
$$
:::

::: fragment
We reject $H_0:\mu=50$ at conventional significance levels.
:::

::: fragment
The sample provides evidence that the population mean differs from 50 hours; the estimate is about 2 hours higher.
:::

------------------------------------------------------------------------

## Common interpretation errors

::: fragment
- “The parameter has a 95% probability of being inside this realized frequentist interval.”
:::

::: fragment
- “A p-value of 0.03 means the null has a 3% probability of being true.”
:::

::: fragment
- “Failure to reject proves the null.”
:::

::: fragment
- “A significant association is causal.”
:::

::: fragment
- “A narrow confidence interval guarantees the study is unbiased.”
:::

::: fragment
Each statement confuses sampling uncertainty with a different inferential question.
:::

------------------------------------------------------------------------

## Exit ticket

::: fragment
1. What is the difference between a parameter and an estimator?
:::

::: fragment
2. What is the difference between a standard deviation and a standard error?
:::

::: fragment
3. What does the LLN tell us? What does the CLT add?
:::

::: fragment
4. Give a correct one-sentence interpretation of a p-value.
:::

::: fragment
5. Name one problem that a larger sample does not automatically solve.
:::

------------------------------------------------------------------------

## Keep this mental map

$$
\boxed{
\begin{array}{c}
\text{Probability}=\text{model of uncertainty}\\[3pt]
\text{Estimator}=\text{random rule based on a sample}\\[3pt]
\text{LLN}=\text{closeness to the target}\\[3pt]
\text{CLT}=\text{shape of estimation error}\\[3pt]
\text{SE}=\text{sampling uncertainty}\\[3pt]
\text{Inference}=\text{estimate + uncertainty + assumptions}
\end{array}}
$$

------------------------------------------------------------------------

**Questions?**

Primary reference: Hansen, *Probability and Statistics for Economists*, Chapters 1-8, 10-11, and 13-14.
